Title here

### 2020-07-27 paper/paper.Rmd
fix notation as per meeting 18

Derivative free methods provides another option to optimise the objective function without evaluating any gradient or Hessian information and a particular class of methods: direct search, has gained its popularity through its conceptual simplicity.  However, the whole searching process in the algorithm remains a black-box. Plots are usually used to evaluate and compare the performance of different algorithms but it can easily become tedious because the code will have to be modified significantly when comparing different parameters in the algorithms. For example, a categorical variable with 5 levels can be easily mapped onto color while mapping another categorical variable with 30 levels will not make the plot informative. Thus the plot needs to be re-designed to better suits the characteristics of the parameter (whether the parameter is a scalar or a matrix? whether the parameter is quantitative or categorical? If categorical, how many levels does the parameter have?). This motivates the  design of a visual diagnostic framework for optimisation algorithms based on the idea of a *global object*.

@@

The optimisation problem we're interested in is in the context of projection pursuit. Coined by @friedman1974projection, projection pursuit is a method that detects the interesting structure (i.e. clustering, outliers and skewness) of multivariate data via projecting it in lower dimensions. Let $\mathbf{X}_{n \times p}$ be a data matrix, an n-d projection can be seen as a linear transformation $T: \mathbb{R}^p \mapsto \mathbb{R}^d$ defined by $\mathbf{P} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{P}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection basis. Define $f: \mathbb{R}^{p \times d} \mapsto \mathbb{R}$ be an index function that maps the projection basis $\mathbf{A}$ onto a scalar value $I$, this function is commonly known as the projection pursuit index (PPI) function, or the index function and can be used to measure the "interestingness" of the projection. A number of indice functions have been proposed in the literature including Legendre index [@friedman1974projection], hermite index [@hall1989polynomial], natural hermite index [@cook1993projection], chi-square index [@posse1995projection], LDA index [@lee2005projection] and PDA index [@lee2010projection].

@@

Projection pursuit is usually used in conjunction with a tour method called *guided tour*. Tour explores the multivariate data *interactively* via playing a series of projections, that form a *tour path* and guided tour uses the path that is geodesically the shortest. Details of the mathematical construction of a tour path can be found in @buja2005computational.  Guided tour, along with other types of tour, has been implemented in the *tourr* package in R, available on the Comprehensive R Archive Network at [https://cran.r-project.org/web/packages/tourr/](https://cran.r-project.org/web/packages/tourr/) [@wickham2011tourrpackage].

@@

## Notation

For a projection basis, let the three-subscript notation $$\mathbf{A}_{jlk}$$ represent the projection matrix in iteration $j$ with either a searching index $l$ or an interpolation index $k$. The placeholder $*$ will be the substitue for the non-applicable index. To give two examples, $\mathbf{A}_{12*}$ denotes the second searching basis in the first iteration; $\mathbf{A}_{1*2}$ denotes the second interpolation basis in the first iteration.

If we are being pedantic, all the target bases, except the starting basis, should have a searching index since it is the last candidate basis that successfully allows the optimisation algorithm to end and output a new basis for interpolation. If we denotes the length of each search (number of basis searched in each iteration) as $l_1, l_2, \cdots, l_J$ and the length of each interpolation is $k_1, k_2, \cdots, k_J$ , then the searching index for target basis in each iteration will become $\mathbf{A}_{1l_1*}, \mathbf{A}_{2l_2*}, \cdots, \mathbf{A}_{Jl_J*}$. We simplify this notation for target bases by using a superscript as $$\mathbf{A}^{0}, \mathbf{A}^{1}, \mathbf{A}^{2}, \cdots, \mathbf{A}^{J}$$ where $\mathbf{A}^{0}$ is the starting basis. Notice that in the case when the complex notation can't be avoided, we shall separate each index using a comma and  write the index in brackets if necessary. Using the notation above, all the bases on the interpolation path (including target and interpolation bases) can be written as in Equation \ref{eq:interp-path}.

\begin{equation}
\{ 
\mathbf{A}^0, 
\mathbf{A}^1, 
\mathbf{A}_{1 \ast 1}, 
\mathbf{A}_{1 \ast 2},
\cdots, 
\mathbf{A}^2, 
\mathbf{A}_{2 \ast 1}, 
\mathbf{A}_{2 \ast 2}, 
\cdots, 
\mathbf{A}^J,
\mathbf{A}_{J \ast 1}, 
\mathbf{A}_{J \ast 2}, 
\cdots, 
\mathbf{A}_{J \ast K},
\}
\label{eq:interp-path}
\end{equation}

where $K$ denotes the last interpolating index. A visual representation of the interpolation path ( for iteration one) modified from @buja2005computational is shown in Figure \ref{tour-path}. If we use a dot to represent a plane in the searching space, the searching points along with the current and target basis in iteration one can sketched in Figure \ref{searching-points}. Combine the two above, we can write all the bases in the first iteration (including the starting basis) in its natural ordering as in Equation \ref{eq:vectorA}.

\begin{equation}
\{
\mathbf{A}^{0}, 
\mathbf{A}_{1 1\ast}, 
\mathbf{A}_{1 2\ast}, 
\cdots, 
\mathbf{A}^{1},
\mathbf{A}_{1\ast 1}, 
\mathbf{A}_{1\ast 2},  
\cdots, 
\mathbf{A}_{1\ast k_1}
\}
\label{eq:vectorA}
\end{equation}

@@

```{r searching-points, out.width="100%", out.height="60%", fig.cap="\\label{searching-points}An illustration of the tour path"}
knitr::include_graphics(path = here::here("figures/tour_path_increasing.png"))
```

@@

\begin{align}
&\arg \max_{\mathbf{A} \in \mathcal{A}} f(\mathbf{A}) \\
s.t. & \mathcal{A} = \{ \forall \mathbf{a}_i, \mathbf{a}_j \in \mathbf{A}: \lVert \mathbf{a}_i \rVert = 1 \text{ and } \mathbf{a}_i \mathbf{a}_j = 0 \}
\end{align}

@@

There are three existing methods for optimisating PPI function and we review them below.

@posse1995projection proposed a stochastic direct search method, a random search algorithm. Given a current basis $\mathbf{A}^{j-1}$, a candidate basis $\mathbf{A}_{jl*}$ is sampled in the neighbourhood defined by the radius of the p-dimensional sphere, $\alpha$ of the current basis $\mathbf{A}^{j}$ by $\mathbf{A}_{jl*} = \mathbf{A}^{j-1} + \alpha \mathbf{A}_{rand}$. If the candidate basis has a higher index value than the current basis, it is outputed as the target basis $\mathbf{A}^{j}$ along with other metadata and the current iteration stops. If no basis is found to have higher index value after the maximum number of tries $l_{\max}$, the algorithm stops with nothing outputted. $c$ is the halfing parameter and when $c > 30$, the nieghbourhood parameter $\alpha$ in the next iteration will be reduced by half. The algorithm for a random search is summarised in Algorithm \ref{random-search}.

\begin{algorithm}
\SetAlgoLined
  \SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}^{j-1}$; The index function: $f$}
    \Output{The global object; The target basis: $\mathbf{A}^{j}$}
  initialisation\;
  \While{$l < l_{\max}$}{
    Generate $\mathbf{A}_{jl*} = \mathbf{A}^{j-1} + \alpha \mathbf{A}_{\text{rand}}$ ensuring $\mathbf{A}_{j, l,*}$ is orthonormal\;
    Compute $I_{jl*}  = f(\mathbf{A}_{jl*})$\;
    \eIf{$I_{jl*} > I^{j-1}$}{
      $A^{j} = A_{jl*}$, $I^{j} = I_{jl*}$\;
      }{
      $c = c + 1$\;
      }
    $l = l + 1$\;
  }
  \caption{random search}
  \label{random-search}
\end{algorithm}

@@

\begin{algorithm}
\SetAlgoLined
\SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}^{j-1}$; The index function: $f$}
    \Output{The global object; The target basis: $\mathbf{A}^{j}$}
  initialisation\;
  \While{$l < l_{\max}$}{
    Generate ten bases in five random directions: $\mathbf{A}_{jl*}: \mathbf{A}_{j,(l + 9), *}$ within a small neighbourhood $\text{dist}$\;
    Find the direction with the largest index value: $\mathbf{A}_{j, l_{d}, *}$\;
    Construct the geodesic from $\mathbf{A}^{j-1}$ to $\mathbf{A}_{j, l_{d}, *}$\;
    Find $\mathbf{A}_{j,l_g, *}$ on the geodesic that has the largest index value \;
    Compute $I_{j,l_g, *} = f(\mathbf{A}_{j,l_g, *})$, $p_{\text{diff}} = (I_{j,l_g, *} - I^{j-1})/I_{j,l_g, *}$\;
      \If{$p_{\text{diff}} > 0.001$}{
        $A^{j} = A_{j,l_g, *}$, $I^{j} = I_{j,l_g, *}$\;
      }
    $l = l + 1$\;
  }
  \caption{search geodesic}
  \label{search-geodesic}
\end{algorithm}

Simulated annealing [@bertsimas1993simulated, @kirkpatrick1983optimization] is a non-derivative procedure based on a non-increasing cooling scheme $T(l)$. Given an initial $T_0$, the temperature at iteration $l$ is defined as $T(l) = \frac{T_0}{log(l + 1)}$. The simulated annealing algorithm works as follows. Given a neighbourhood parameter $\alpha$ and a randomly generated orthonormal basis $A_{rand}$, a candidate basis is constructed as $\mathbf{A}_{jl*} = (1 - \alpha)\mathbf{A}^{j} + \alpha \mathbf{A}_{rand}$. If the index value of the candidate basis is larger than the one of the current basis, the candidate basis becomes the target basis. If it is smaller, the candidate is accepted with probability $P= \min\left\{\exp\left[\frac{I^{j-1} - I_{jl*}}{T(l)}\right],1\right\}$ where $I^{j-1}$ and $I_{jl*}$ are the index value of the current and candidate basis respectively. The algorithm can be summarised as in Algorithm \ref{simulated_annealing}.

\begin{algorithm}
\SetAlgoLined
\SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}^{j-1}$; The index function: $f$}
    \Output{The global object; The target basis: $\mathbf{A}^{j}$}
  initialisation\;
  \While{$l < l_{\max}$}{
    Generate $\mathbf{A}_{jl*} = (1 - \alpha)\mathbf{A}^{j-1} + \alpha \mathbf{A}_{rand}$ ensuring $\mathbf{A}_{jl*}$ is orthonormal \;
    Commpute $I_{jl*} = f(\mathbf{A}_{jl*})$ and $T(l) = \frac{T_0}{\log(l + 1)}$\;
      \eIf{$I_{jl*} > I^{j-1}$}{
        $A^{j} = A_{jl*}$, $I^{j} = I_{jl*}$\;
      }{
        Compute $P= \min\left\{\exp\left[\frac{I^{j-1} - I_{jl*}}{T(l)}\right],1\right\}$\;
        Draw $D$ from a uniform distribution: $D \sim \text{Unif(0, 1)}$\;
        \If{$P > D$}{
          $A^{j} = A_{jl*}$, $I^{j} = I_{jl*}$\;
        }
      }
    $l = l + 1$\;
  }
  \caption{simulated annealing}
  \label{simulated_annealing}
\end{algorithm}

@@

Ggplot requires a data frame that contains all the variables to plot and a *global object* is constructed as the data frame supplied to the visual diagnostic plots to better suit the characters of iterative optimisation algorithms. Given an optimisation algorithm, two primary variables of interest are the *decision variable: $\mathbf{A}$* and the *value of the objective function: $I$*. To further simplify the notation for $\mathbf{A}$, we write the bases as a column vector denoting by a single subscript by their natural ordering as in Equation \ref{vector-basis}.

\begin{equation}
\{
\mathbf{A}_0, 
\mathbf{A}_1, 
\mathbf{A}_2, 
\cdots,
\mathbf{A}_T
\label{vector-basis}
\}
\end{equation}

where $T$ is the total number of plane serached by the optimisation algorithm including all the target planes, all the interpolating planes and all the candidate planes but NOT the starting plane.

*Iterators* indexes the data collected and has a time series feature that prescribes the its natural ordering of the decision varible in the searching. The simpliest iterator, *index per observation: $t$*, is a unique identifier for each observation in the data. For each level of iteration, we design two types of indices: *index per iteration: $j$* is fixed for each observations in one iteration and has an increment of one once a new iteration starts. *index within iteration: $l$* has an increment of one for each observation in an iteartion and starts over from one once a new iteration starts. A sketch of the difference between these three iterators is provided in Figure \ref{iterators}. In projection pursuit optimisation algorithms, there is one level of iteartion and hence exists three iterators: `id` indices each observation by a unique number; `tries` is the index per iteration iterator that gets updated once a search-and-interpolate step is finished; and `loop` is the index within iteration iterator and starts over from one at the beginning of a new `tries`. We give the interpolating basis a different index $k$ for projection pursuit guided tour. It is similar in nature to $l$ but is specific for interpolating bases, which is usually not part of the optimisation.

```{r iterators, out.width = "100%", out.height = "25%", fig.cap = "\\label{iterators} A sketch of the design of iterators in iterative algorithms."}
knitr::include_graphics(here::here("figures", "iterators.png"))
```

There could exist other parameters that are also of our interest but can't be classified as one of the three categories. They are defined under the fourth category: *other parameter of interest: $S$*. In projection pursuit, three other parameters of our interest includes `method`, `alpha` and `info`.  `method` identifies the name of the searching method used and we are interested in comparing the performance between different algorithms under direct search. The neighbourhood parameter `alpha` controls the size of the sampling space and we are interested to understand how the searching space shrinks as the algorithm progresses. `info` labels different stages in the searching process. A sketch of the global object for projection pursuit guided tour is presented in Figure \ref{fig:glb-obj}. The full data structure can thus be shown as in Equation \ref{eq:data-structure}.

\begin{equation}
\left[
\begin{array}{c|cc|ccc|ccc}
t & \mathbf{A} & I & j &  l  & k & S_{1}& \cdots & S_{p}\\
\hline
1 & \mathbf{A}_0 & I_0 & 0 & \ast & \ast & \ast & \cdots & \ast\\
\hline
2 & \mathbf{A}_1 & I_1 & 1 & 1 & \ast & S_{1, 1} & \cdots & S_{1, p}\\
3 & \mathbf{A}_2 & I_2 & 1 & 2 & \ast & S_{2, 1} & \cdots & S_{2, p}\\
\vdots & \vdots &\vdots & \vdots &\vdots & \vdots &\vdots &\cdots &\vdots\\
\vdots & \vdots & \vdots & 1 & l_1 & \ast & \vdots &  \cdots & \vdots\\
\hline
\vdots &\vdots & \vdots & 1 & \ast & 1& \vdots &  \cdots & \vdots\\
\vdots &\vdots &\vdots& 1 & \ast & 2& \vdots& \cdots & \vdots\\
\vdots &\vdots &\vdots &\vdots  &\vdots & \vdots &\vdots &\cdots &\vdots \\
\vdots &\vdots &\vdots & 1 & \ast & k_1&\vdots & \cdots & \vdots\\
\hline
\vdots &\vdots &\vdots &\vdots  &\vdots & \vdots &\vdots &\cdots &\vdots \\
T+1 & \mathbf{A}_T & I_T & J &  l_J & \ast & S_{T, 1}& \cdots & S_{T, p}\\
\end{array}
\right]
\label{eq:data-structure}
\end{equation}

@@

We first simulate some random variables of size 1000 with different structures. `x1`, `x8`, `x9` and `x10` are simulated from normal distribution iwth zero mean and variance of one as in equation \ref{eq:sim-norm}. When using projection pursuit to explore the data structure based on its departure from normality, the entry in the projection basis for these variables should be close to zero in theory. `x2` to `x7` are mixture of normal distribution with different weights and locations. Equation \ref{eq:sim-x2} to \ref{eq:sim-x7} show the distribution of each variables and Figure \ref{origin-data} shows the histogram of each variable except `x3`. All the variables are then scaled to ensure the mixture has variance of one.

@@

In the first two sections, we will first provide a toy example that is easy to grasp and then more examples that can help us to understand the algorithm and parameter choice. Remeber the research question we raised earlier, the purpose of visual diagnostics is to understand:

@@

The first question can be answered using a static plot with x-axis showing the progression of the optimisation and y-axis showing the value of the objective function. The second question can be addressed via visualising the rotating high dimensional space or its projection on the reduced 2D space. Thus animated visualisation is needed to preceive the optimisation path in the searching space.

@@

- Layer 1: boxplot geom
  - data: group by `tries` and filter the observations in the group that have count greater than `cutoff = 15`.
  - x: `tries` is mapped to the x-axis
  - y: the index value after statistical transformation $y = \{Q_q(I_{jlk} \cdot \mathds{1} (j = x, k = \ast))\}$ is mapped to the y-axis where $q$ takes one of $0, 25, 50, 75, 100$ and $Q_q(x)$ finds the qth-quantile for the vector $x$. $\mathds{1}(.)$ is the identity operator.

- Layer 2: point geom
  - data: group by `tries` and filter the observations in the group that have count less than `cutoff = 15`.
  - x: `tries` is mapped to the x-axis
  - y: `index_val` is mapped to the y-axis

- Layer 3: line geom
  - data: filter the points with the highest index value in each `tries`
  - x: `tries` is mapped to the x-axis
  - y: `index_val` is mapped to the y-axis

@@

- Layer 1: point geom
  - data: filter the observations with `info` being `interpolation` and mutate `id` to be the row number of the subsetted tibble
  - x: `id` is mapped to the x-axis
  - y: `index_val` is mapped to the y-axis

- Layer 2: line geom
  - using line goemetry for the same data and aesthetics

@@

<!-- In the previous two sections, only the iterator and the index value are mapped onto the x and y aesthetics of the plot; while more aesthetics i.e. color, could be added to compare other parameters in the global object. -->

@@

- Layer 1: point geom
  - data: filter the observations with `info` being `interpolation`; bind the global object from optimisation and interpolation and form polishing; mutate `id` to be the row number of the binded tibble.
  - x: `id` is mapped to the x-axis
  - y: `index_val` is mapped to the y-axis 
  - color: `method` is mapped to the color aesthetic

- Layer 2: line geom
  - using line goemetry for the same data and aesthetics

@@

Again using the same data, we are interested to compare the effect of different `max.tries` in the 2D projection setting. `max.tries` is a hyperparameter that controls the maximum number of try before the search ends. The default value of 25 is suitable for 1D projection while we suspect it may not be a good option for the 2D case and hence want to compare it with an alternative, 500. As shown in Figure \ref{trace-compare}, both trials attain the same index value after polishing while the small `max.tries` of 25 is not sufficient for `search_better` to find its global maximum and we will need to adjust the `max.tries` argument for the search to succiciently explore the parameter space.

@@

<!-- The resulting two vectors are of dimension $n \times 1$, which is compatible with the dimension of the global object. THis means all the iterators and interesting other variables defined in the global object can be binded with the these two projections and can be mapped to other aesthetics (i.e. color). -->

- Layer 1: point geom
  - data: subset the basis of interest and arrange into a matrix format; perform PCA on the basis matrix and compute the projected basis on the first two principal components; bind the variables from the original global object and form a tibble
  - x: the projected basis on the first principal component
  - y: the projected basis on the second principal component
  - color: an variable of interest is mapped onto color

@@

```{r pca, fig.cap = "\\label{pca}PCA plot of search geodesic Coloring by info allows for better understanding of each stage in the geodesic search"}

@@

Polishing step has done much more work to find the final index value ub `search_geodesic` than `search_better` and this indicates `kol_cdf` function favours of a random search method than ascent method.

>>>>

Derivative free methods provides another option to optimise the objective function without evaluating any gradient or Hessian information and a particular class of methods: direct search, has gained its popularity through its conceptual simplicity.  However, the whole searching process in the algorithm remains a black-box. Plots are usually used to evaluate and compare the performance of different algorithms but it can easily become tedious because the code will have to be modified significantly when comparing different parameters in the algorithms. For example, a categorical variable with 5 levels can be easily mapped onto colour while mapping another categorical variable with 30 levels will not make the plot informative. Thus the plot needs to be re-designed to better suits the characteristics of the parameter (whether the parameter is a scalar or a matrix? whether the parameter is quantitative or categorical? If categorical, how many levels does the parameter have?). This motivates the  design of a visual diagnostic framework for optimisation algorithms based on the idea of a *global object*.

@@

The optimisation problem we're interested in is in the context of projection pursuit. Coined by @friedman1974projection, projection pursuit is a method that detects the interesting structure (i.e. clustering, outliers and skewness) of multivariate data via projecting it in lower dimensions. Let $\mathbf{X}_{n \times p}$ be a data matrix, an n-d projection can be seen as a linear transformation $T: \mathbb{R}^p \mapsto \mathbb{R}^d$ defined by $\mathbf{P} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{P}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection basis. Define $f: \mathbb{R}^{p \times d} \mapsto \mathbb{R}$ be an index function that maps the projection basis $\mathbf{A}$ onto a scalar value $I$, this function is commonly known as the projection pursuit index (PPI) function, or the index function and can be used to measure the "interestingness" of the projection. A number of index functions have been proposed in the literature including Legendre index [@friedman1974projection], Hermite index [@hall1989polynomial], natural Hermite index [@cook1993projection], chi-square index [@posse1995projection], LDA index [@lee2005projection] and PDA index [@lee2010projection].

@@

Projection pursuit is usually used in conjunction with a tour method called *guided tour*. Tour explores the multivariate data *interactively* via playing a series of projections, that form a *tour path* and guided tour uses the path that is geodesically the shortest. Details of the mathematical construction of a tour path can be found in @buja2005computational.  [Maybe add the sketch for illustrating tour path.] Guided tour, along with other types of tour, has been implemented in the *tourr* package in R, available on the Comprehensive R Archive Network at [https://cran.r-project.org/web/packages/tourr/](https://cran.r-project.org/web/packages/tourr/) [@wickham2011tourrpackage].

@@

@@

@@

\begin{align}
&\arg \max_{\mathbf{A} \in \mathcal{A}} f(\mathbf{X} \cdot \mathbf{A}) \\
s.t. & \mathcal{A} = \{ \forall \mathbf{a}_i, \mathbf{a}_j \in \mathbf{A}: \lVert \mathbf{a}_i \rVert = 1 \text{ and } \mathbf{a}_i \mathbf{a}_j = 0 \}
\end{align}

@@

There are three existing methods for optimising PPI function and we review them below.

@posse1995projection proposed a stochastic direct search method, a random search algorithm. Given a current basis $\mathbf{A}_{\text{cur}}$, a candidate basis $\mathbf{A}_{t}$ is sampled in the neighbourhood defined by the radius of the p-dimensional sphere, $\alpha$ of the current basis $\mathbf{A}_{t}$ by $\mathbf{A}_{t} = \mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{rand}$. If the candidate basis has a higher index value than the current basis, it is outputted as the target basis $\mathbf{A}_{t}$ along with other metadata and the current iteration stops. If no basis is found to have higher index value after the maximum number of tries $l_{\max}$, the algorithm stops with nothing outputted. $c$ is the halfing parameter and when $c > 30$, the neighbourhood parameter $\alpha$ in the next iteration will be reduced by half. The algorithm for a random search is summarised in Algorithm \ref{random-search}.

\begin{algorithm}
\SetAlgoLined
  \SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}_{\text{cur}}$; The index function: $f$}
    \Output{The target basis: $\mathbf{A}_{t+1}$}
  initialisation\;
  Set $l = 1$\;
  \While{$l < l_{\max}$}{
    Generate $\mathbf{A}_{l} = \mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}$ ensuring $\mathbf{A}_{l}$ is orthonormal\;
    Compute $I_{l}  = f(\mathbf{A}_{l})$\;
    \eIf{$I_{l} > I_{\text{cur}}$}{
      \KwRet{$\mathbf{A}_{l}$} \;
      }{
      $c = c + 1$\;
      }
    $l = l + 1$\;
  }
  \caption{random search}
  \label{random-search}
\end{algorithm}

@@

\begin{algorithm}
\SetAlgoLined
\SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}_{\text{cur}}$; The index function: $f$}
    \Output{The target basis: $\mathbf{A}_{**}$}
  initialisation\;
  Set $l = 1$\;
  \While{$l < l_{\max}$}{
    Generate ten bases in five random directions: $\mathbf{A}_{l}: \mathbf{A}_{l+9}$ within a small neighbourhood $\text{dist}$\;
    Find the direction with the largest index value: $\mathbf{A}_{*}$ where $l < * < l+ 9$\;
    Construct the geodesic $\mathcal{G}$ from $\mathbf{A}_{\text{cur}}$ to $\mathbf{A}_{*}$\;
    Find $\mathbf{A}_{**}$ on the geodesic $\mathcal{G}$ that has the largest index value \;
    Compute $I_{**} = f(\mathbf{A}_{**})$, $p_{\text{diff}} = (I_{**} - I_{\text{cur}})/I_{**}$\;
      \If{$p_{\text{diff}} > 0.001$}{
         \KwRet{$\mathbf{A}_{**}$} \;
      }
    $l = l + 1$\;
  }
  \caption{search geodesic}
  \label{search-geodesic}
\end{algorithm}

Simulated annealing [@bertsimas1993simulated, @kirkpatrick1983optimization] is a non-derivative procedure based on a non-increasing cooling scheme $T(t)$. Given an initial $T_0$, the temperature at iteration $t$ is defined as $T(t) = \frac{T_0}{log(t + 1)}$. The simulated annealing algorithm works as follows. Given a neighbourhood parameter $\alpha$ and a randomly generated orthonormal basis $A_{rand}$, a candidate basis is constructed as $\mathbf{A}_{t+1} = (1 - \alpha)\mathbf{A}_{t} + \alpha \mathbf{A}_{rand}$. If the index value of the candidate basis is larger than the one of the current basis, the candidate basis becomes the target basis. If it is smaller, the candidate is accepted with probability $P= \min\left\{\exp\left[\frac{I_{t+1} - I_{t}}{T(t)}\right],1\right\}$ where $I_{t}$ and $I_{t+1}$ are the index value of the current and candidate basis respectively. The algorithm can be summarised as in Algorithm \ref{simulated_annealing}.

\begin{algorithm}
\SetAlgoLined
\SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}_{\text{cur}}$; The index function: $f$}
    \Output{The global object; The target basis: $\mathbf{A}^{j}$}
  initialisation\;
  \While{$l < l_{\max}$}{
    Set $l = 0$\;
    Generate $\mathbf{A}_{l} = (1 - \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{rand}$ ensuring $\mathbf{A}_{l}$ is orthonormal \;
    Compute $I_{l} = f(\mathbf{A}_{l})$ and $T(l) = \frac{T_0}{\log(l + 1)}$\;
      \eIf{$I_{l} > I_{\text{cur}}$}{
        \KwRet{$\mathbf{A}_{l}$} \;
      }{
        Compute $P= \min\left\{\exp\left[\frac{I_{\text{cur}} -I_{l}}{T(l)}\right],1\right\}$\;
        Draw $D$ from a uniform distribution: $D \sim \text{Unif(0, 1)}$\;
        \If{$P > D$}{
           \KwRet{$\mathbf{A}_{l}$} \;
        }
      }
    $l = l + 1$\;
  }
  \caption{simulated annealing}
  \label{simulated_annealing}
\end{algorithm}

@@

Ggplot requires a data frame that contains all the variables to plot and a *global object* is constructed as the data frame supplied to the visual diagnostic plots to better suit the characters of iterative optimisation algorithms. Given an optimisation algorithm, two primary variables of interest are the *decision variable: $\mathbf{A}$* and the *value of the objective function: $I$*, which is the projection basis and index value repsectively in projection pursuit. *State: $S$* is another variable of great importance since it labels $\mathbf{A}$ and $I$ by the stage they are in the whole search-and-interpolate process.

*Iterators* indexes the data collected and prescribes the the natural ordering of each observation in different levels. *$t$* is a unique identifier that has an increment of one for each observation and *$j$* is the counter for each search-and-interpolate round. [add a bit more for *$l$* after decide whether to combine $l$ and $k$. ]

<!-- For each level of iteration, we design two types of indices: *index per iteration: $j$* is fixed for each observations in one iteration and has an increment of one once a new iteration starts. *index within iteration: $l$* has an increment of one for each observation in an iteration and starts over from one once a new iteration starts. -->

<!-- In projection pursuit optimisation algorithms, there is one level of iteration and hence exists three iterators: `id` indices each observation by a unique number; `tries` is the index per iteration iterator that gets updated once a search-and-interpolate step is finished; and `loop` is the index within iteration iterator and starts over from one at the beginning of a new `tries`. We give the interpolating basis a different index $k$ for projection pursuit guided tour. It is similar in nature to $l$ but is specific for interpolating bases, which is usually not part of the optimisation. -->

<!-- # ```{r iterators, out.width = "100%", out.height = "25%", fig.cap = "\\label{iterators} A sketch of the design of iterators in iterative algorithms."} -->
<!-- # knitr::include_graphics(here::here("figures", "iterators.png")) -->
<!-- # ``` -->

There are other parameters that are of our interest and we denote them as *$V_{p}$*. In projection pursuit, this includes $V_1 = \text{method}$, denoting the name of the searching method used and $V_2 = \text{alpha}$, which is a neighbourhood parameter that controls the size in sampling candidate bases. A sketch of the global object for projection pursuit guided tour is presented in Figure \ref{fig:glb-obj}. The full data structure can thus be shown as in Equation \ref{eq:data-structure}.

\begin{equation}
\left[
\begin{array}{c|ccc|ccc|ccc}
t & \mathbf{A} & I & S & j &  l  & k & V_{1}& \cdots & V_{p}\\
\hline
1 & \mathbf{A}_1 & I_1 & S_1 & 1 & 1 & \ast & V_{1, 1} & \cdots & V_{p, 1}\\
\hline
2 & \mathbf{A}_2 & I_2 & S_2 & 2 & 1 & \ast & V_{1, 2} & \cdots & V_{p, 2}\\
3 & \mathbf{A}_3 & I_3 & S_3 & 2 & 2 & \ast & V_{1, 3} & \cdots & V_{p, 3}\\
\vdots & \vdots &\vdots &\vdots & \vdots &\vdots & \vdots &\vdots &\cdots &\vdots\\
\vdots & \vdots & \vdots &\vdots & 2 & l_1 & \ast & \vdots &  \cdots & \vdots\\
\hline
\vdots &\vdots & \vdots &\vdots & 2 & \ast & 1& \vdots &  \cdots & \vdots\\
\vdots &\vdots &\vdots &\vdots & 2 & \ast & 2& \vdots& \cdots & \vdots\\
\vdots &\vdots &\vdots &\vdots &\vdots  &\vdots & \vdots &\vdots &\cdots &\vdots \\
\vdots &\vdots &\vdots &\vdots & 2 & \ast & k_1&\vdots & \cdots & \vdots\\
\hline
\vdots &\vdots &\vdots &\vdots &\vdots  &\vdots & \vdots &\vdots &\cdots &\vdots \\
T & \mathbf{A}_T & I_T &S_T  & J &  l_J & \ast & V_{1, T}& \cdots & V_{p, T}\\
\end{array}
\right]
\label{eq:data-structure}
\end{equation}

@@

We first simulate some random variables of size 1000 with different structures. `x1`, `x8`, `x9` and `x10` are simulated from normal distribution with zero mean and variance of one as in equation \ref{eq:sim-norm}. When using projection pursuit to explore the data structure based on its departure from normality, the entry in the projection basis for these variables should be close to zero in theory. `x2` to `x7` are mixture of normal distribution with different weights and locations. Equation \ref{eq:sim-x2} to \ref{eq:sim-x7} show the distribution of each variables and Figure \ref{origin-data} shows the histogram of each variable except `x3`. All the variables are then scaled to ensure the mixture has variance of one.

@@

In the first two sections, we will first provide a toy example that is easy to grasp and then more examples that can help us to understand the algorithm and parameter choice. Remember the research question we raised earlier, the purpose of visual diagnostics is to understand:

@@

The first question can be answered using a static plot with x-axis showing the progression of the optimisation and y-axis showing the value of the objective function. The second question can be addressed via visualising the rotating high dimensional space or its projection on the reduced 2D space. Thus animated visualisation is needed to perceive the optimisation path in the searching space.

@@

- Layer 1: boxplot geom
  - data: group by $j$ and filter the observations in the group that have count greater than `cutoff = 15`.
  - x: $j$ is mapped to the x-axis
  - y: the statistical transformed index value: $Q_q(I_{T} \cdot \mathds{1} (S = \text{search}))$ is mapped to the y-axis where $q$ takes one of $0, 25, 50, 75, 100$ and $Q_q(x)$ finds the qth-quantile for the vector $x$. $\mathds{1}(.)$ is the identity operator.

- Layer 2: point geom
  - data: group by $j$ and filter the observations in the group that have count less than `cutoff = 15`.
  - x: $j$ is mapped to the x-axis
  - y: $I$ is mapped to the y-axis

- Layer 3: line geom
  - data: filter the points with the highest index value in group $j$
  - x: $j$ is mapped to the x-axis
  - y: $I$ is mapped to the y-axis

@@

- Layer 1: point geom
  - data: filter the observations with $S = \text{interpolation}$ and mutate $t$ to be the row number of the subsetted tibble
  - x: $t$ is mapped to the x-axis
  - y: $I$ is mapped to the y-axis

- Layer 2: line geom
  - using line geometry for the same data and aesthetics

@@

<!-- In the previous two sections, only the iterator and the index value are mapped onto the x and y aesthetics of the plot; while more aesthetics i.e. colour, could be added to compare other parameters in the global object. -->

@@

- Layer 1: point geom
  - data: filter the observations with  $S = \text{interpolation}$; bind the global object from optimisation and interpolation and form polishing; mutate $t$ to be the row number of the binded tibble.
  - x: $t$ is mapped to the x-axis
  - y: $I$ is mapped to the y-axis 
  - colour: $V$ is mapped to the colour aesthetic

- Layer 2: line geom

@@

Again using the same data, we are interested to compare the effect of different `max.tries` in the 2D projection setting. `max.tries` is a hyperparameter that controls the maximum number of try before the search ends. The default value of 25 is suitable for 1D projection while we suspect it may not be a good option for the 2D case and hence want to compare it with an alternative, 500. As shown in Figure \ref{trace-compare}, both trials attain the same index value after polishing while the small `max.tries` of 25 is not sufficient for `search_better` to find its global maximum and we will need to adjust the `max.tries` argument for the search to sufficiently explore the parameter space.

@@

<!-- The resulting two vectors are of dimension $n \times 1$, which is compatible with the dimension of the global object. THis means all the iterators and interesting other variables defined in the global object can be bound with the these two projections and can be mapped to other aesthetics (i.e. colour). -->

- Layer 1: point geom
  - data: subset the basis of interest and arrange into a matrix format; perform PCA on the basis matrix and compute the projected basis on the first two principal components; bind the variables from the original global object and form a tibble
  - x: the projected basis on the first principal component
  - y: the projected basis on the second principal component
  - colour: $V$ is mapped to the colour aesthetic

@@

```{r pca, fig.cap = "\\label{pca}PCA plot of search geodesic colouring by info allows for better understanding of each stage in the geodesic search"}

@@

Polishing step has done much more work to find the final index value in `search_geodesic` than `search_better` and this indicates `kol_cdf` function favours of a random search method than ascent method.
